---
title: "Fitting one-phase and two-phase exponential decays"
author: "XF"
date: "1/11/2022"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install package
# install.packages(c("tidyverse", "nls.multstart", "broom"))
library(tidyverse)
library(broom)
library(nls.multstart)

```

# 1. Fit data with a one-phase exponential model  
Y = span * e^(-k * t) + plateau

`span` is the difference between the start point and plateau in y axis  
`k` is the rate constant  
`plateau` is the Y value at infinite times  

## 1.1 Simulate data  

```{r simulate data}
data_1 <-  data.frame(t = seq(100),
                      Y = 10 * exp(-0.07 * seq(100)) + 5 + rnorm(100, mean = 0, sd = 0.1))
```

## 1.2 Use `nls` by guessing the single starting value  

```{r}
# Select an approximate span, must be lower than min(A), and greater than zero
plateau_guess <- min(data_1$Y) * 0.5

# Estimate the rest parameters using a linear model
model_guess <- lm(log(Y - plateau_guess) ~ t, data = data_1)
span_guess <- exp(coef(model_guess)[1])
k_guess <- -coef(model_guess)[2]

# Starting parameters for the linear model
start <-
  list(span = span_guess,
       k = k_guess,
       plateau = plateau_guess)

# Fit the model (with estimated starting parameters)
fit_1 <-
  nls(Y ~ span * exp(-k * t) + plateau,
      data = data_1,
      start = start)

fit_1
```

### Get fit parameters

```{r}
tidy(fit_1)
```

### Plot fit
```{r}
# Use broom’s augment function to get predicted data
preds <- augment(fit_1)
ggplot() +
  geom_point(aes(t, Y), data_1) +
  geom_line(aes(t, .fitted), preds, colour = "orange", size = 1)
```

## 1.3 Use `nls.multstart`: a gridstart approach with multiple starting parameters 

This method uses a random-search/shotgun approach to fit multiple curves. Random start parameter values are picked from a uniform distribution between start_lower() and start_upper() for each parameter. If the best model is not improved upon (in terms of AIC score) for 100 new start parameter combinations, the function will return that model fit.  

Reference: https://github.com/cran/nls.multstart  

### Run nls_multstart with gridstart approach  

```{r nls_multstart}
fit_1 <- nls_multstart(
          Y ~ span * exp(-k * t) + plateau,
          data = data_1,
          iter = 100,
          start_lower = c(span = 10, k = 0.001, plateau = 4),
          start_upper = c(span = 15, k = 0.01, plateau = 8),
          supp_errors = 'Y',
          lower = c(span = 0, k = 0.001, platea = 0)
        )

fit_1
```

### Get fit parameters  
```{r}
tidy(fit_1)
```

### Plot fit  
```{r}
preds <- augment(fit_1)
ggplot() +
  geom_point(aes(t, Y), data_1) +
  geom_line(aes(t, .fitted), preds, colour = "orange", size = 1)
```

# 2. Fit data with a two-phase exponential model
Y = span\_fast * e^(-k\_fast * t) + span\_slow * e^(-k\_slow * t) + plateau  

`k_fast` and `k_slow` are the two rate constants  
`span_fast` and `span_slow` are the differences between the start points (either fast component or slow component) and plateau in y axis  
`plateau` is the Y value at infinite times  

## 2.1 Simulate data
```{r}
data_2 <-  data.frame(t = seq(100),
                      Y = 14 * exp(-0.5 * seq(100)) +
                      1 * exp(-0.05 * seq(100)) +
                      4 + rnorm(100, mean = 0, sd = 0.1))
```

## Use `nls.multstart`: a gridstart approach with multiple starting parameters 
```{r}
fit_2 <- nls_multstart(Y ~ alpha * exp(-beta * t) + gamma * exp(-delta * t) + theta,
                       data = data_2,
                       iter = 100,
                       start_lower = c(alpha=10, beta=0.01, gamma=0, delta=0.01, theta=4),
                       start_upper = c(alpha=15, beta=1, gamma=5, delta=1, theta=8),
                       supp_errors = 'Y',
                       lower = c(alpha=0, beta=0.001, gamma=0, delta=0.001, theta=0))
fit_2
```

If you run this step multiple times, you will notice that the order of the term with alpha and beta and the term with gamma and delta can be interchanged because of the ramdom starting parameters. You may want to set the term with a larger decay constant as span_fast and k_fast as below.


### Get fit parameters
```{r}
## set the term with a larger decay constant as span_fast and k_fast
if (coef(fit_2)['beta'] > coef(fit_2)['delta']) {
  span_fast <- coef(fit_2)['alpha']
  k_fast <- coef(fit_2)['beta']
  span_slow <- coef(fit_2)['gamma']
  k_slow <- coef(fit_2)['delta']
} else {
  span_fast <- coef(fit_2)['gamma']
  k_fast <- coef(fit_2)['delta']
  span_slow <- coef(fit_2)['alpha']
  k_slow <- coef(fit_2)['beta']
}
plateau <- coef(fit_2)['theta']

names(span_fast) <- 'span_fast'
names(k_fast) <- 'k_fast'
names(span_slow) <- 'span_slow'
names(k_slow) <- 'k_slow'
names(plateau) <- 'plateau'

# Get the sum of the squared residuals.
ssr <- sum(resid(fit_2) ^ 2)
names(ssr) <- 'ssr'

# Return all coefficients
coefficients <- c(span_fast, k_fast, span_slow, k_slow, plateau, ssr)
coefficients
```

## 2. Comparing one-phase and two-phase exponential models

```{r}
# Fit data_2 with a one-phase exponential model
fit_1 <- nls_multstart(
          Y ~ span * exp(-k * t) + plateau,
          data = data_2,
          iter = 100,
          start_lower = c(span = 10, k = 0.001, plateau = 4),
          start_upper = c(span = 15, k = 0.01, plateau = 8),
          supp_errors = 'Y',
          lower = c(span = 0, k = 0.001, platea = 0)
        )

# Compare two models with anova()
anova(fit_1, fit_2)
```
The result shows a Df of 2 (indicating that the more complex model has two additional parameters), and the p-value is small(< 0.001). This means that adding the two additional parameters to the two-phase exponential model did lead to a significantly improved fit over the one-phase exponential model.

### Plot fits
```{r}
preds_1 <- augment(fit_1)
preds_2 <- augment(fit_2)
ggplot() +
  geom_point(aes(t, Y), data_2) +
  geom_line(aes(t, .fitted, colour="one-phase"), preds_1, size= 1)+
  geom_line(aes(t, .fitted, colour="two-phase"), preds_2, size= 1)+
  scale_color_manual(name = "", values = c("one-phase" = "red", "two-phase" = "orange"))

```
